Exploring how changes in cause of death have narrowed the

### 2017-02-10 Manuscript/Manuscript_COD_decomp.Rmd
wrote half of abstract

---
title: Exploring how changes in age at death and cause of death have narrowed the
  black-white life expectancy gap over forty-five years in the United States.
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure are differences by race. Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in the cause of death [*and age at death(?)*] over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013. Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be examined.

@@

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC. In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` in `r sum3[dim(sum3)[1]-1, ]$state ` when DC is excluded). By 2013, it ranged from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` in `r sum4[dim(sum4)[1]-1, ]$state` when DC is excluded). During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardivascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

>>>>

---
title: Exploring how changes in cause of death have narrowed the
  black-white life expectancy gap over forty-five years in the United States.
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

**Background**

The difference in life expectancy between black and whites in the US has decreased over time. States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York undergoing greater reductions to the gap compared to states such as California and Texas. The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in relatively narrow life expectancy gaps in some states and  wide gaps in other states.

**Methods and Findings**
Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes). We excluded 10 states with the smallest black populations, but included the District of Columbia. A bayesian autoregressive model was used to model smoothed mortality rates over time within each strata, and a truncated Poisson model was used to imputed death counts that were suppressed due to small cell sizes. We calculated life expectancy for blacks and whites within each strata using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.

[Add findings]

**Conclusions**
[Add conclusion]

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure are differences by race. Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013. Thus, common reasons for the decreasing life expectancy gap can be identified, and anomalous states can be examined.

@@

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Figure 1c**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` in `r sum3[dim(sum3)[1]-1, ]$state ` when DC is excluded). By 2013, it ranged from `r round(sum4[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum4[dim(sum4)[1], ]$state ` (or `r round(sum4[dim(sum4)[1]-1, ]$LE_wbgap, 1)` in `r sum4[dim(sum4)[1]-1, ]$state` when DC is excluded). During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardivascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.
